---
title: "factors affecting model performance"
output: html_notebook
---

#****************************************
# MIT License
# Copyright (c) 2026 Jinying Chen
#  
# author(s): Jinying Chen, Boston University Chobanian & Avedisian School of Medicine
# release date: 2026-8-15
# ver: 1.0
# 
# This code was written to support data analysis for the
# 2026 paper published in JMI Medical Informatics.
# The code is for research use only, and is provided as it is.
# 

```{r}
library(tidyverse)
library(dplyr)
library(glue)
library(purrr)
library(psych)
```

```{r}
setwd("./")

indir <- "./results/"

rxnorm_infile <- glue("{indir}/rxnorm_test_predictions.csv")
ltc_infile <- glue("{indir}/LTCDC_test_predictions.csv")
  
df_rxnorm <-read_csv(rxnorm_infile)

df_ltc <- read_csv(ltc_infile)
```
# prepocessing

## create length variable for RxNorm
```{r}
df_rxnorm$length <- nchar(df_rxnorm$term)

table(df_rxnorm$type_norm)

df_rxnorm$type_norm <- factor(df_rxnorm$type_norm, levels = c("IN", "BN"))


describeBy(df_rxnorm$length, group = df_rxnorm$type_norm)
```

## create length variable for LTC
```{r}
df_ltc$length <- nchar(df_ltc$term)

table(df_ltc$term_type)

df_ltc$type_norm <- factor(df_ltc$term_type, levels = c("generic", "brand"))

describeBy(df_ltc$length, group = df_ltc$type_norm)
```

## distribution of length variables
```{r}
hist(df_rxnorm$length)
```

```{r}
hist(df_ltc$length)
```


```{r}
colnames(df_rxnorm)
```


# analysis on RxNorm terms
## auxilliary funtion
```{r}
analyze_RxNorm_terms <- function(
  df,
  model_name,
  predictors,
  interaction_terms = NULL,
  conf_level = 0.95
) {
  pred_col <- paste0(model_name, "__pred")
  correct_col <- paste0(model_name, "__correct")
  
  if (!pred_col %in% names(df)) {
    stop("Prediction column not found: ", pred_col)
  }
  if (!"target" %in% names(df)) {
    stop("Column `target` not found.")
  }
  
  if (length(predictors) == 0) {
    stop("At least one predictor must be provided.")
  }
  
  missing_predictors <- setdiff(predictors, names(df))
  if (length(missing_predictors) > 0) {
    stop(
      "Predictor column(s) not found: ",
      paste(missing_predictors, collapse = ", ")
    )
  }
  
  # Standardize interaction_terms
  # Examples:
  # interaction_terms = c("A+B", "term length")
  # interaction_terms = list(c("A+B", "term length"), c("has_digit", "edit distance"))
  if (!is.null(interaction_terms)) {
    if (is.character(interaction_terms)) {
      interaction_terms <- list(interaction_terms)
    }
    
    if (!is.list(interaction_terms)) {
      stop(
        "`interaction_terms` must be NULL, a character vector, ",
        "or a list of character vectors."
      )
    }
    
    for (x in interaction_terms) {
      if (!is.character(x) || length(x) < 2) {
        stop("Each interaction term must contain at least two variable names.")
      }
    }
    
    interaction_vars <- unique(unlist(interaction_terms, use.names = FALSE))
    
    missing_interaction_vars <- setdiff(interaction_vars, names(df))
    if (length(missing_interaction_vars) > 0) {
      stop(
        "Interaction variable(s) not found: ",
        paste(missing_interaction_vars, collapse = ", ")
      )
    }
  } else {
    interaction_vars <- character(0)
  }
  
  # Create the model-specific correct column
  df[[correct_col]] <- as.integer(df[[pred_col]] == df[["target"]])
  
  # Work on a modeling copy so we do not return temporary safe-name columns
  model_df <- df
  
  # Include predictors and variables used in interaction terms
  all_model_vars <- unique(c(predictors, interaction_vars))
  
  # Create safe temporary names for formula construction
  safe_predictor_names <- setNames(
    paste0(".rxnorm_x", seq_along(all_model_vars)),
    all_model_vars
  )
  
  # Add safe-name copies to model_df
  for (v in all_model_vars) {
    model_df[[safe_predictor_names[[v]]]] <- model_df[[v]]
  }
  
  # Temporary safe outcome name
  tmp_outcome <- ".correct_outcome_tmp"
  model_df[[tmp_outcome]] <- model_df[[correct_col]]
  
  # Main-effect terms not already included through interaction terms
  if (!is.null(interaction_terms)) {
    interaction_vars <- unique(unlist(interaction_terms, use.names = FALSE))
    
    other_predictors <- setdiff(predictors, interaction_vars)
    other_terms <- unname(safe_predictor_names[other_predictors])
    
    interaction_formula_terms <- vapply(
      interaction_terms,
      function(vars) {
        safe_vars <- unname(safe_predictor_names[vars])
        
        # Using * includes both main effects and the interaction.
        # For example: A * B = A + B + A:B
        paste(safe_vars, collapse = " * ")
      },
      character(1)
    )
    
    final_terms <- c(other_terms, interaction_formula_terms)
  } else {
    final_terms <- unname(safe_predictor_names[predictors])
  }
  
  form <- as.formula(
    paste(tmp_outcome, "~", paste(final_terms, collapse = " + "))
  )
  
  fit <- glm(
    formula = form,
    data = model_df,
    family = binomial()
  )
  
  s <- summary(fit)$coefficients
  
  alpha <- 1 - conf_level
  z_value <- qnorm(1 - alpha / 2)
  
  coef_table <- data.frame(
    model = model_name,
    term = rownames(s),
    estimate = s[, "Estimate"],
    std_error = s[, "Std. Error"],
    z_value = s[, "z value"],
    p_value = s[, "Pr(>|z|)"],
    odds_ratio = exp(s[, "Estimate"]),
    ci_lower = exp(s[, "Estimate"] - z_value * s[, "Std. Error"]),
    ci_upper = exp(s[, "Estimate"] + z_value * s[, "Std. Error"]),
    row.names = NULL
  )
  
  # Convert safe temporary names back to original names
  pretty_term <- function(x) {
    ord <- order(nchar(unname(safe_predictor_names)), decreasing = TRUE)
    
    for (i in ord) {
      safe_name <- unname(safe_predictor_names[i])
      original_name <- names(safe_predictor_names)[i]
      x <- gsub(safe_name, original_name, x, fixed = TRUE)
    }
    
    x
  }
  
  coef_table$term <- vapply(coef_table$term, pretty_term, character(1))
  
  list(
    data = df,
    model_name = model_name,
    pred_col = pred_col,
    correct_col = correct_col,
    predictors = predictors,
    interaction_terms = interaction_terms,
    formula_used = form,
    predictor_name_map = safe_predictor_names,
    fit = fit,
    summary = summary(fit),
    coef_table = coef_table
  )
}
```



```{r}
model_names <- sub("__pred$", "", grep("__pred$", names(df_rxnorm), value = TRUE))

predictors <- c("length", "type_norm")
```


## multiple regression
```{r}
all_results1 <- lapply(model_names, function(m) {
  analyze_RxNorm_terms(
    df = df_rxnorm,
    model_name = m,
    predictors = predictors
  )
})

names(all_results1) <- model_names
```


```{r}
all_results1$BERTDrug$coef_table
```


```{r}
all_results1$CharBERTDrug$coef_table
```


# analysis on LTC terms

```{r}
colnames(df_ltc)

table(df_ltc$data_type)
```
## auxilliary function
```{r}
analyze_ltc_terms <- function(
  df,
  model_name,
  predictors,
  threshold = 0.5,
  interaction = NULL
) {
  prob_col <- paste0(model_name, "_probability")
  correct_col <- paste0(model_name, "_correct")
  
  if (!prob_col %in% names(df)) {
    stop("Probability column not found: ", prob_col)
  }
  if (!"target" %in% names(df)) {
    stop("Column `target` not found.")
  }
  
  if (!is.null(interaction)) {
    if (length(interaction) != 2) {
      stop("`interaction` must be NULL or a character vector of length 2.")
    }
    
    predictors <- unique(c(predictors, interaction))
  }
  
  missing_predictors <- setdiff(predictors, names(df))
  if (length(missing_predictors) > 0) {
    stop(
      "Predictor column(s) not found: ",
      paste(missing_predictors, collapse = ", ")
    )
  }
  
  predicted_class <- ifelse(df[[prob_col]] >= threshold, 1, 0)
  df[[correct_col]] <- as.integer(predicted_class == df[["target"]])
  
  tmp_outcome <- ".y_correct"
  safe_predictor_names <- paste0(".p", seq_along(predictors))
  names(safe_predictor_names) <- predictors
  
  model_df <- data.frame(
    .y_correct = df[[correct_col]]
  )
  
  for (p in predictors) {
    model_df[[safe_predictor_names[[p]]]] <- df[[p]]
  }
  
  if (!is.null(interaction)) {
    safe_interaction <- safe_predictor_names[interaction]
    
    other_predictors <- setdiff(predictors, interaction)
    other_terms <- unname(safe_predictor_names[other_predictors])
    
    interaction_term <- paste(unname(safe_interaction), collapse = " * ")
    final_terms <- c(other_terms, interaction_term)
  } else {
    final_terms <- unname(safe_predictor_names[predictors])
  }
  
  form <- as.formula(
    paste(tmp_outcome, "~", paste(final_terms, collapse = " + "))
  )
  
  fit <- glm(
    formula = form,
    data = model_df,
    family = binomial()
  )
  
  s <- summary(fit)$coefficients
  
  coef_table <- data.frame(
    model = model_name,
    term = rownames(s),
    estimate = s[, "Estimate"],
    std_error = s[, "Std. Error"],
    z_value = s[, "z value"],
    p_value = s[, "Pr(>|z|)"],
    odds_ratio = exp(s[, "Estimate"]),
    ci_lower = exp(s[, "Estimate"] - 1.96 * s[, "Std. Error"]),
    ci_upper = exp(s[, "Estimate"] + 1.96 * s[, "Std. Error"]),
    row.names = NULL
  )
  
  pretty_term <- function(x) {
    ord <- order(nchar(unname(safe_predictor_names)), decreasing = TRUE)
    
    for (i in ord) {
      safe_name <- unname(safe_predictor_names[i])
      original_name <- names(safe_predictor_names)[i]
      x <- gsub(safe_name, original_name, x, fixed = TRUE)
    }
    
    x
  }
  
  coef_table$term <- vapply(coef_table$term, pretty_term, character(1))
  
  list(
    data = df,
    model_name = model_name,
    prob_col = prob_col,
    correct_col = correct_col,
    predictors = predictors,
    interaction = interaction,
    threshold = threshold,
    formula_used = form,
    predictor_name_map = safe_predictor_names,
    fit = fit,
    summary = summary(fit),
    coef_table = coef_table
  )
}
```

```{r}
model_names <- sub("_probability$", "", grep("_probability$", names(df_ltc), value = TRUE))

predictors <- c("length", "type_norm")
```

## OOV terms, multiple regression
```{r}
all_results_ltc_oov1 <- lapply(model_names, function(m) {
  analyze_ltc_terms(
    df = df_ltc[df_ltc$data_type == "oov", ],
    model_name = m,
    predictors = predictors  
    )
})

names(all_results_ltc_oov1) <- model_names
```

```{r}
all_results_ltc_oov1$BERTDrug$coef_table
```

```{r}
all_results_ltc_oov1$CharBERTDrug$coef_table
```
